---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# intervalaverage

<!-- badges: start -->
  [![Travis build status](https://travis-ci.com/kaufman-lab/intervalaverage.svg?branch=main)](https://travis-ci.com/kaufman-lab/intervalaverage)
<!-- badges: end -->


Perform fast and memory efficient time-weighted averaging of values 
    measured over intervals into new arbitrary intervals. 
    
    
This package is useful in the context of data measured or represented
as constant values over intervals on a one-dimensional discrete axis 
(e.g. time-integrated averages of a curve over defined periods). 
	
	
The intervalaverage package was written specifically to deal with air 
pollution data recorded or predicted as averages over sampling periods. 
Data in this format often needs to be shifted to non-aligned periods 
or averaged up to longer durations 	(e.g. averaging data measured over 
sequential non-overlapping periods to calendar years). 
	
	
This package makes careful use of optimized data.table syntax to reduce copying 
so it is well suited to deal with large datasets. The function inputs and return 
values are always data.tables (with values, interval starts
and interval ends stored as columns) which provide a straightforward 
representation of the data facilitating direct inspection and manipulation via
data.table's standard interface. Functions also accept optional grouping variables as
arguments; these group-by operations get passed to data.table which avoids the overhead
of looping directly in R.
	

## Installation

Right now, the only way to install the package is from [GitHub](https://github.com/):

``` r
# install.packages("devtools")
devtools::install_github("kaufman-lab/intervalaverage","main",build_vignettes=TRUE)
```

## Example

```{r}
library(intervalaverage)
```

Consider some PM2.5 data measured over periods occuring over a repeating 7-day schedule:
```{r}
(x <- data.table(start=seq(1L,by=7L,length=6),
                 end=seq(7L,by=7L,length=6),
                 pm25=c(10,12,8,14,22,18)))
```

Imagine that we would like these PM2.5 measurements to be represented over a different set of intervals, such as the following 7-day schedule which does not align cleanly with the intervals in x:

```{r}
(y <- data.table(start=seq(3L,by=7L,length=6),
                 end=seq(9L,by=7L,length=6)))
```

Representing the PM25 measurements over the exact intervals in y is not
strictly possible, since each y interval contains two
x intervals which only partially overlap.

While there are several ways to compromise and approximate a representation of x over the periods in y, one reasonable approach is to average all the values of PM2.5 in x occuring over the interval of y. But since the two overlap durations of the two x periods into a single y period are unequal, the PM2.5 values should be averaged in a way that takes the duration of overlap into account. I.e, we need a time-weighted average of values of x into intervals in y.

This is the exact purpose of the intervalaverage function:


```{r}
(z <- intervalaverage(x,y,interval_vars=c("start","end"),
                value_vars=c("pm25")))
```

The pm25 value of 10.571429 is equal to `(5/7)*10 + (2/7)*12`:

```{r}
identical(z[1,pm25], (5/7)*10 + (2/7)*12)
```


Note that in z the average value of PM2.5 is missing (NA) for the final period in y.  This is because there is insufficient PM25 to calculate an average.  By using a looser competeness requirements we can get an average for this period in y (in general, this will just be the mean of non-missing values in x):

```{r}
(z <- intervalaverage(x,y,interval_vars=c("start","end"),
                value_vars=c("pm25"),required_percentage = 70))
```

There is much more described in the vignette. For example, it is possible to calculate averages within a grouping variable. For example if PM2.5 were measured at multiple locations, you could take the average separately at each location in one call to the `intervalaverage` function. Averaging intervals need not be the same for each locations. See the vignette for examples.


## Vignette

Read the intro vignette for an in-depth demonstration of the package functions.

```
library(intervalaverage)
vignette("intervalaverage-intro")
```

For an overview of the internals of the  `intervalaverage` function, see the technical vignette:

```
vignette("intervalaverage-technicaloverview")
```


