---
title: "Introduction to intervalaverage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intervalaverage-intro}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo=FALSE,message=FALSE}
library(intervalaverage)
set.seed(1)
```


This package and vignette makes extensive use of `data.table`. If you're unfamiliar with the 
`data.table` syntax, a brief review of that package's introductory vignette may be useful.


## Averaging values measured over intervals


Consider the following dataset which represents average (predicted) pm2.5 exposure  and no2 exposure
at some location over four sequential 7-day periods at the beginning of the year 2000:
```{r}
exposure_dataset <- data.table(location_id=1, start=seq(as.IDate("2000-01-01"),by=7,length=4),
                end=seq(as.IDate("2000-01-07"),by=7,length=4),pm25=rnorm(4,mean=15), no2=rnorm(4,mean=25))
exposure_dataset
```
Note that the above intervals are stored as a column for the start of the interval and a column for the end of the interval.  For the purpose of this package, intervals are ALWAYS treated as closed (i.e. inclusive of start and end values) and the variables storing interval starts and ends must be discrete (e.g., class integer or IDate).


If we wanted to calculate the average of the first two weeks of pm25 data, this would simply be the
average of the two pm25 values for those weeks:
```{r}
exposure_dataset[start %in% as.IDate(c("2000-01-01","2000-01-08")),mean(pm25)]
```

But we wanted the average of the first 10 days of that pm25 data, we would need to take a weighted average 
since the period from Jan 1 to Jan 10 doesn't align perfectly with the intervals over which the pm2.5
data is recored:
```{r}
exposure_dataset[start %in% as.IDate(c("2000-01-01","2000-01-08")),weighted.mean(pm25,w=c(7/10,3/10))]
```

The `intervalaverage` package and specifically the `intervalaverage` function was written
to facilitate this sort averaging operation.  In order to use this the package function, we'll need a
dataset containing data that's stored over intervals (such as in `exposure_dataset`) as well as a dataset containing
the periods you'd like to average over. 

Let's create a dataset containing some periods we'd like average over:

```{r}
averaging_periods <- data.table(start=seq(as.IDate("2000-01-01"),by=10,length=3),
                end=seq(as.IDate("2000-01-10"),by=10,length=3))
averaging_periods
```


Now that we have defined intervals to average over, let's use the `intervalaverage` function to calculate
the averages:

Note that in order for the intervalaverage function to work, the start and end columns need to have
the same column names in x and in y. These column names are specified via the `interval_vars` 
argument. And the variables in `exposure_dataset` that you want averages calculated for are specified 
via `value_vars`.

```{r}
averaged_exposures <- intervalaverage(exposure_dataset,averaging_periods,                                      interval_vars=c("start","end"),value_vars=c("pm25","no2"))
averaged_exposures[, list(start,end,pm25,no2)]
```
The return value of the `intervalaverage` function is a `data.table`. Above I'm just displaying the first four columns of that return object.  Note that the value of the pm25 in the first row is what we calculated manually above, and the `start` and `end` columns in that row are indicating that that pm25
value is for the period from Jan 1, 2000 to Jan 10, 2000.

Also note that the function allowed us to calculate averages for multiple variables simulateously by 
specifying a vector of column names in the `value_vars` argument. Finally, note that the third entry for
both the `pm25` and the `no2` column is `NA` or missing. This makes sense because the original dataset
`exposure_dataset` didn't have data for every day in the interval from Jan 21, 2000 to Jan 30, 2000.

Displaying the full data.table returned by the function gives us some more infomration:
```{r}
averaged_exposures
```
The `xduration` column tells us the number of days that were even present in `exposure_dataset` for each interval specified in `averaging_periods`. 
The first two `averaging_periods` intervals were fully represented in `exposure_dataset`, whereas `exposure_dataset` only contained data for 
8 of the 10 days in the third `averaging_periods` interval.  The `xmaxend` column shows us that the last day in the interval from Jan 21,2000 to
Jan 30, 2000 that was present in `exposure_dataset` was Jan 28, 2000.

These supplementary columns are useful for diagnosing incomplete data in `exposure_dataset`.

If we're ok with calculating an average based on incomplete data, we can set the the tolerance for 
missingness lower. Let's say we're ok with calculating an average if 75% or more of the period is observed:
```{r}
intervalaverage(exposure_dataset,averaging_periods,interval_vars=c("start","end"),
                value_vars=c("pm25","no2"),required_percentage = 75)
```
The results are the same for the first two rows but now we have nonmissing values in the third
which are calculated based on the available data in `exposure_dataset`.
If there had been a period with less than 75% of the data present, the function would still return `NA`
for those value variables.



## Averaging within values of a grouping variable (e.g. at multiple locations)
Typically we're dealing with more than one location at a time. Let's create a data.table similar to 
`exposure_dataset` but with several (three) locations:

```{r}
exposure_dataset2 <- rbindlist(lapply(1:3, function(z){
  data.table(location_id=z, 
             start=seq(as.IDate("2000-01-01"),by=7,length=4),
             end=seq(as.IDate("2000-01-07"),by=7,length=4),pm25=rnorm(4,mean=15),
             no2=rnorm(4,mean=25))} ))
exposure_dataset2
```

If we want to calculate separate averages of values in `exposure_dataset2` over the periods in `averaging_periods`
for each of the `location_id`s in `exposure_dataset`, we need a version of the `averaging_periods` table
that contains every interval for every possible `location_id`.  The reason why the
function is written this way will be come clear in the a subsequent section (where different averaging periods
are used for each location).  For now, let's say we want the same averaging periods for every location.

```{r}
#unexpanded:
averaging_periods
#expanded to every location_id:
rbindlist(lapply(1:3, function(z)copy(averaging_periods)[,location_id:=z][])) 
```

The above code is a bit esoteric so I've written a function to simplify and generalize this process
of repeating/expanding a set of intervals (or more generally, a set of rows in a table) 
for every `location_id`  (or more generally, for every row in another table). To use this `CJ.dt` function, 
just create a `data.table` with a column containing unique ids, then call `CJ.dt` on the two tables:

```{r}
exposure_dataset2_unique_locs <- data.table(location_id=unique(exposure_dataset2$location_id))
averaging_periods2 <- CJ.dt(averaging_periods, exposure_dataset2_unique_locs)
averaging_periods2
```


Now, all we have to do is use the same call as above to intervalaverage while specifying one more argument: `group_vars="location_id"`.
```{r}
intervalaverage(exposure_dataset2,averaging_periods2,interval_vars=c("start","end"),
                value_vars=c("pm25","no2"),
                group_vars="location_id",
                required_percentage = 75)[, list(location_id, start,end, pm25,no2)]
```
Of course, we could have completed the above by calling intervalaverage repeatedly for each value of `location_id` 
in x an y using a for loop. The reason to prefer using the `group_vars` approach is that
the `intervalaverage` function is written to be faster than looping when with dealing with grouping.
It also saves you the trouble of writing a loop and combining the results.





## Values in overlapping periods are not allowed

Note that all the intervals used in this package are treated as inclusive.  So far, we've dealt
with data which have intervals which do not overlap.  However, consider the following dataset where
the end day of a previous interval is the start day of the next interval:
```{r}
exposure_dataset_overlapping <- rbindlist(lapply(1:3, function(z){
  data.table(location_id=z, start=seq(as.IDate("2000-01-01"),by=7,length=4),
             end=seq(as.IDate("2000-01-08"),by=7,length=4),pm25=rnorm(4,mean=15),
             no2=rnorm(4,mean=25) )
} ))
exposure_dataset_overlapping
```

If we try to average this exposure dataset, we get an errror:
```{r,error=TRUE}
intervalaverage(exposure_dataset_overlapping,averaging_periods2,
                interval_vars=c("start","end"),
                value_vars=c("pm25","no2"),
                group_vars="location_id",
                required_percentage = 75)
```
That's because the `intervalaverage` function is written to throw an error if there are overlaps in within groups. This
is to encourage the user to explicitly and conciously deal with overlaps prior to averaging.

Note that we can also check whether there are overlapping intervals using `is.overlapping`
```{r}
is.overlapping(exposure_dataset_overlapping, interval_vars=c('start','end'),group_vars="location_id")
```

In order to deal with partially overlapping intervals, we need to split intervals into areas of exact overlap and non-overlap with the `isolateoverlaps` function:

```{r}
exposure_dataset_isolated <- isolateoverlaps(exposure_dataset_overlapping, interval_vars=c("start","end"), 
                                             group_vars="location_id",
                                             interval_vars_out=c("start2","end2"))
exposure_dataset_isolated[1:15] #only show the first 15 rows
```
Inspect the above table and compare it to `exposure_dataset`. `start2` and `end2` are the new intervals
and `start` and `end` are the original intervals. Note how there are two rows for every
overlapping period (ie in the `start2` and `end2` columnds), but the pm25 and no2 values differ 
within these rows since one value comes from the first overlapping period and the second value comes 
from the second overlapping period

We can then average exposure values within periods of exact overlap:
```{r}
exposure_dataset_overlaps_averaged <-
  exposure_dataset_isolated[, list(pm25=mean(pm25),no2=mean(no2)),by=c("location_id","start2","end2")]
setnames(exposure_dataset_overlaps_averaged, c("start2","end2"),c("start","end"))
```

This version of the dataset where values in overlapping periods have already been averaged can 
now be averaged to a the times in `averaging_periods2` using the `intervalaverage` function:

```{r}
intervalaverage(exposure_dataset_overlaps_averaged,averaging_periods2,interval_vars=c("start","end"),value_vars=c("pm25","no2"),
                group_vars="location_id",
                required_percentage = 75)[,list(location_id, start,end,pm25,no2)]
```

## Averaging values measured over intervals: different averaging periods for diferent observations
Often we're interested in calculating averages over different periods for different locations.
First to make this more realistic, let's generate ~20 years of data at 2000 locations:

```{r}
exposure_dataset3 <- rbindlist(lapply(1:2000, function(z){
  data.table(location_id=z, start=seq(as.IDate("2000-01-01"),by=7,length=1000),
                end=seq(as.IDate("2000-01-07"),by=7,length=1000),pm25=rnorm(4,mean=15),
                no2=rnorm(4,mean=25) )
  } ))
```

Now let's pick a different random end date for each location's averaging period. We'll define the start
date as 3 years before that to define three-year intervals for every location.
```{r}
averaging_periods3 <- data.table(location_id=1:2000,
                                 end=sample(
                                   seq(as.IDate("2001-01-01"),as.IDate("2019-01-01"),by=1),
                                   2000
                                 )
)
averaging_periods3[,start:=end-round(3*365.25)]
averaging_periods3
```

Calculating the average is even easier than in the previous example since we don't need to "expand" 
y to all the location_ids (since it's been generated to have different start and end dates for each location):
```{r}
intervalaverage(exposure_dataset3,averaging_periods3,interval_vars=c("start","end"),value_vars=c("pm25","no2"),
                group_vars="location_id",
                required_percentage = 75)[,list(location_id,start,end,pm25,no2)]
```
If you were to inspect this table, you shouldn't be surprised to see some missingness since the 
earliest possible averaging interval end is Jan 1 2001 and the first observed value of the pollutants is Jan 1 2000.


Finally, a quick trick if you'd like to calculate 1-year, 2-year, and 3-year averages all at once:
```{r}
averaging_periods3[, avg3yr:=end-round(3*365.25)]
averaging_periods3[, avg2yr:=end-round(2*365.25)]
averaging_periods3[, avg1yr:=end-round(1*365.25)]
#reshape the data.table:
averaging_periods4 <- melt(averaging_periods3,id.vars=c("location_id","end"),
                           measure.vars = c("avg3yr","avg2yr","avg1yr")) 
setnames(averaging_periods4, "value","start")
setnames(averaging_periods4, "variable","averaging_period")
averaging_periods4
```

```{r}

intervalaverage(exposure_dataset3,averaging_periods4,interval_vars=c("start","end"),
                value_vars=c("pm25","no2"),
                group_vars=c("location_id"),
                required_percentage = 75)[,list(location_id,start,end,pm25,no2)]
```



## Interval intersects: averaging with an address history
So far we've fully covered the functionality of the `intervalaverage` function and how to use it
when we want to average over time at specific locations. 

The above examples also cover the approach we'd use if we wanted
to average over a cohort of study participants for whom we only have a single address (and we are 
ok assuming that participants never move).

However, in cohort studies we often have a set of locations/addresses for each participant and time intervals
indicating when the participant lived at which address. Typically this information is represented through 
a table we refer to as an "address history."

We'll start with a very simple example to demonstrate what an address history looks like 
and how we might use this in exposure averaging. Consider the following address history and exposure datasets:

```{r,echo=FALSE}
address_history0 <- data.table(addr_id=c(1,2,2,3,5),
                 ppt_id=c(1,1,1,2,2),
                 addr_start=c(1L,10L,12L,1L,13L),
                 addr_end=c(9L,11L,14L,12L,15L)) 
exposure_dataset5 <- data.table(addr_id=rep(1:4,each=3),
  exposure_start=rep(c(1L,8L,15L),times=4),
  exposure_end=rep(c(7L,14L,21L),times=4),
  exposure_value=c(rnorm(12))
 )
```
```{r}
exposure_dataset5
```

Note that `exposure_dataset5` has two regular (length-7) intervals for each address and corresponding measurements for those periods.

```{r}
address_history0
```
The first thing to note is that the address intervals (`addr_start` and `addr_end`) are non-overlapping,
which is good because `intervalaverage` requires non-overlapping intervals as we saw previously

The address_history0 has one participant with three rows and two addresses (`addr_id`s 1 and 2).
In practice this would be the same data if the two intervals where `ppt_id==1 & addr_id==2` were
stored as a single row corresponding to the interval `[10,14]`, but I've created the dataset like
this to demonstrate that a single address represented over non-overlapping intervals doesn't cause problems.
The second participant also has two addresses (`addr_ids`s 3 and 5).

The goal here is to get exposures merged to addresses, but the problem is that the address periods 
don't line up nicely with the exposure periods. Participant 1 lived add address 1 from `[1,9]` 
but exposure is measured over `[1,7]` and `[8,14]`.  The solution is to create two rows for that 
participant, one row for `[1,7]` and a second row from `[8,9]`. This can be accomplished using the 
`intervalintersect` function:


```{r}

 exposure_addresss_table <- intervalintersect(exposure_dataset5,
                                              address_history0,
                interval_vars=c(exposure_start="addr_start",exposure_end="addr_end"),
                "addr_id")
```
`intervalintersect` takes every possible combination of overlapping intervals within `group_vars`
(in this sense it is a cartesian join. More on this below). 

`intervalintersect` is also an inner beacuse rows from either table that are not joined are not 
included in the output. For example, rows where `addr_id==4` in `exposure_dataset5` is not included 
since there are no rows in `address_history0` where `addr_id==4`. Additionally, none of the exposure
periods from `exposure_dataset5` measured over `exposure_start==15` to `exposure_start==21` are 
included in the result, because none of the address history intervals overlap with those periods.
Finally, there is an address (`addr_id==5` from `ppt_id==2`) in the `addr_history` table that 
isn't in the `exposure_dataset5` table.  This address is also excluded from the result since 
exposure estimates do not exist for that participant.

In general the `intervalintersect` joins every interval it can. If a row you expected is missing 
from the result, it's because it could not be joined. It's worth doing some checks after
completion of the intersection to identify what information has been dropped:

```{r}
setdiff(address_history0$addr_id,exposure_addresss_table$addr_id)
setdiff(exposure_dataset5$addr_id,exposure_addresss_table$addr_id)
```

Finally, note that the syntax of `interval_vars` allows those columns to be named differently
in `x` and `y` via a named vector: `interval_vars=c(exposure_start="addr_start",exposure_end="addr_end")`.


### Interval intersects: averaging with a larger address history
starting with the unique set of locations extracted from `exposure_dataset3`, let's generate 300 participants and a random number of addresses each participant lived at
```{r}
n_ppt <- 300
addr_history <- data.table(ppt_id=paste0("ppt",1:n_ppt))
addr_history[, n_addr := rbinom(.N,size=length(unique(exposure_dataset3$location_id)),prob=.005)] 

#repl=TRUE because it's possible for an address to be lived at multiple different time intervals:
addr_history <- addr_history[, 
                             list(location_id=sample(exposure_dataset3$location_id,n_addr,replace=TRUE)),
                             by="ppt_id"
                             ]   
addr_history

#note that not all of these 2000 locations in exposure_dataset3 were "lived at" in this cohort:
length(unique(addr_history$location_id)) 

#also note that it's possible for different participants to live at the same address.
addr_history[,list(loc_with_more_than_one_ppt=length(unique(ppt_id))>1),by=location_id][,sum(loc_with_more_than_one_ppt)]
#Because of the way I generated this data, it's way more common than you'd expect in a real cohort 
#but it does happen especially in cohorts with familial recruitment or people living in nursing home complexes.
```

I've generated intervals which are non-overlapping representing the particpant address history
(the code to achieve this is hidden because it's complicated and not the point of this vignette):
```{r,echo=FALSE}
#generate a vector from which dates will be sampled

sample_dates <- function(n){
  stopifnot(n%%2==0)
  dateseq <- seq(as.IDate("1995-01-01"),as.IDate("2017-01-01"),by=1)
  dates <- sort(sample(dateseq,n))
  #half of the time, make the last date "9999-01-01" which represents that the currently
   #lives at that location and we're carrying that assumption forward
  if(runif(1)>.5){
    dates[length(dates)] <- as.IDate("9999-01-01") 
  }
  dates
}


addr_history_dates <- addr_history[,list(date=sample_dates(.N*2)) ,by="ppt_id"] #for every address, ppt needs two dates: start and end
addr_history_dates_wide <- addr_history_dates[, list(start=date[(1:.N)%%2==1],end=date[(1:.N)%%2==0]),by="ppt_id"]
addr_history <- cbind(addr_history,addr_history_dates_wide[, list(start,end)])
setnames(addr_history, c("start","end"),c("addr_start","addr_end"))
#addr_history[,any(end=="9999-01-01"),by="ppt_id"][, sum(V1)]
setkey(addr_history, ppt_id, addr_start)
#here i'm using a trick to map distinct values of location_id to integers (within ppt) by coercing to factor then back to numeric.
addr_history[, addr_id:=paste0(ppt_id, "_", as.numeric(as.factor(location_id))),by=ppt_id]
```

```{r}
addr_history
```
Oftentimes participant addresses are given their own keys that are distinct from location_id 
and that's represented in the above table.  This means that a single `location_id` may map to multiple
`addr_ids`.
 
It's important for these address tables to be non-overlapping within ppt. 
As shown previously, there's a function in the `intervalaverage` package for that check:
```{r}
is.overlapping(addr_history,interval_vars=c("addr_start","addr_end"),
               group_vars="ppt_id") #FALSE is good--it means there's no overlap of dates within ppt.
```

This table passes that check because I've generated the data to be non-overlapping, but often times
people report overlapping address histories and analytic decisions need to be made to de-overlap them.


### interval intersects is a cartesian join
So far we've seen exposure datasets stored by `location_id` but it's possible also to store exposures
stored by `addr_id` (such that the series of exposure estimates for a single locations may
be repeated multiple time if that `location_id` maps to multiple `addr_id`s )
```{r}
exposure_dataset3_addr <- unique(addr_history[, list(location_id,addr_id)])[exposure_dataset3, 
                                                                            on=c("location_id"),
                                                                            allow.cartesian=TRUE,
                                                                            nomatch=NULL]
exposure_dataset3
exposure_dataset3_addr
```

Note that `exposure_dataset3_addr` contains repeat locations whereas `exposure_dataset3` 
contains exactly one location per time point:

```{r}
exposure_dataset3[, sum(duplicated(location_id)),by=c("start")][,max(V1)] #no duplicate locations at any date
exposure_dataset3_addr[, sum(duplicated(location_id)),by=c("start")][,max(V1)] 
```
`exposure_dataset3_addr` has duplicate locations since multiple ppts may live at the same location
or because a single participant lives at the same location multiple times.

Storing exposure data according to `addr_id` rather than `location_id` takes up more space but 
may be beneficial depending on various data pipeline/dataflow processes.

`location_id` may not even be present in the address table if it's stored by address_id:
```{r}
exposure_dataset3_addr[, location_id:=NULL]
```


Even if this disctinction between how exposures are stored doesn't seem relevant, 
this section will demonstrate how `intervalintersect`
is actually a cartesian join (in addition to being an inner interval join).


Whether the exposure dataset is stored by address or location, the `intervalintersect` 
will result in values from the exposure dataset merged to every address. In the case of the 
exposure table being stored by `addr_id`, this is a simple one to one merge (since for every
address in the address history there's a set of exposures in the exposure table).

But in the case of the exposure table being stored by `location_id`, this becomes a one to many merge
(since a single location id may merge to multiple locations in the address history table).

This works because the function that `intervalintersect` relies on (`data.table::foverlaps`)
is performing an inner cartesian merge: that is--it only takes rows which match on the keying variables
but also does a cartesian expansion if there are multiple matches in one or both tables. If exposures are keyed on 
`location_id` then this table needs to be expanded since a single location may be used in
more than one address_history record.

```{r}
z <- intervalintersect(x=exposure_dataset3,
               y=addr_history,
               interval_vars=c(
                 start="addr_start",
                 end="addr_end"
                 ),
               group_vars=c("location_id"),
               interval_vars_out=c("start2","end2")
) 

z_addr <- intervalintersect(x=exposure_dataset3_addr,
               y=addr_history,
               interval_vars=c(
                 start="addr_start",
                 end="addr_end"
                 ),
               group_vars=c("addr_id"),
               interval_vars_out=c("start2","end2")
) 

setkey(z,ppt_id,start2,end2)
setkey(z_addr,ppt_id,start2,end2)
all.equal(z,z_addr)
```
