---
title: "Quick Introduction to intervalaverage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intervalaverage-intro}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE) 
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo=FALSE,message=FALSE}
library(intervalaverage)
```

*Note: This package and vignette makes extensive use of `data.table`. If
you're unfamiliar with the `data.table` syntax, a brief review of that
package's [introductory
vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
may be useful.*

## What does `intervalaverage` do? 

The `intervalaverage` package is intended to do a specific job very
efficiently. Namely, it averages data measured non-continuously over
arbitrary intervals. It is implemented in C++ and `data.table` in order
to do this job as fast and in as memory-efficient a way as possible.

The motivation for this package was to efficiently create time- and
location-weighted averages of air pollution exposure for participants in
epidemiological studies. In these studies, air pollution exposure is
modeled on a weekly, monthly, or annual basis for study participants'
home addresses. In order to create a long-term average of each study
participant's air pollution exposure, it is necessary to calculate an
average of the modeled exposures for each address for the period during
which they lived there, which may not align nicely with the exposure
periods produced by the models.

There are likely many other applications for averaging non-continuous
measurements over arbitrary intervals, but this documentation will
demonstrate the application of calculating long-term air pollution exposures
for individuals over multiple locations. 

## Data

We will use two data sets (included in this package) to demonstrate
`intervalaverage`'s main functionality.

-   `no2` Nitrogen Dioxide (NO~2~) modeled annually at 62 home addresses
    (identified by `location_id`)

-   `address_history` The time periods for which 25 people (identified
    by `person_id`) lived at each of these home address.

```{r}
library(intervalaverage)
data("no2")
data("address_history")
setDT(no2) 
setDT(address_history)
head(no2[])
head(address_history[])
```

Note that each of these data sets is already a `data.table` and the date columns
are `IDate`s. However, due to a quirk in `data.table`, we need to run `setDT` 
on data loaded with the `data` function. 

## Averaging Measurements over Arbitrary Intervals

Inspecting the `no2` data set above, we see that each location has a
single NO~2~ measurement for each calendar year. It would be simple to
find the average measurement over, for example, 2002-01-01 through
2005-12-31, because that period aligns with our measurements. However,
what if we wanted to find the average NO~2~ level at each of the periods
in our `address_history`? Not only does each address have a different
date range to average over, but that range doesn't align nicely with the
measurements! For each address, we need to take a mean of the
measurements, weighted by how much of the measurement was overlapped by
the address history.

This is straightforward with `intervalaverage`'s `intervalaverage`
function!

The `y` argument of `intervalaverage` tells it what date range to
average for every group in the values data set (in this case, every
`location_id` in `no2`). For now, we don't care about who lived at each
address, only the unique addresses and date ranges. Since some people in
our address history shared homes, we begin by creating a set of unique
addresses and their associated dates.

```{r}
unique_addresses <- unique(address_history[ , .(location_id, start_date, end_date)])
```

Now we can use the `intervalaverage` function:

```{r}
averaged_exposures <- intervalaverage(
  x = no2,
  y = unique_addresses,
  interval_vars = c("start_date", "end_date"),
  value_vars = "no2", 
  group_vars = "location_id"
)
head(averaged_exposures[ , .(location_id, start_date, end_date, no2)])
```

You'll notice that there are many missing values. By default,
`intervalaverage` returns `NA` unless the interval specified in `y` is
completely covered by the values in `x`, and many of our addresses were
occupied outside the time period covered by the NO~2~ model. To relax
this requirement and return a value even when not all of the interval
had NO~2~ values, use the `required_percentage` argument. For example,
here we accept an average if at least 80% of the interval has values.

```{r}
averaged_exposures <- intervalaverage(
  x = no2,
  y = unique_addresses,
  interval_vars = c("start_date", "end_date"),
  value_vars = "no2", 
  group_vars = "location_id", 
  required_percentage = 80
)
head(averaged_exposures[ , .(location_id, start_date, end_date, no2)])
```

`intervalaverage` also includes some diagnostic columns in its output
that can be useful in determining why `NA`s are being returned (this is
why we selected only `location_id`, `start_date`, `end_date`, and `no2`
to display above). See the "Value" section of `help(intervalaverage)`
for an explanation of what these columns mean.

## Averaging over an Address History

If you look closely at our address history, you'll see that some
locations and dates are shared among multiple `person_id`s. This makes
sense, since more than one person often lives in the same house! You'll
also notice that people move (so each `person_id` may be associated with
multiple `location_id`s). 

To create an average for each person (rather than location), we'll need
to link the `no2` measurement at each location to the person who lived
there at the time. Remember, multiple people might have lived at the
same location at the same or different times.

The `intervalaverage` package does this with the `intervalintersect`
function. It looks very similar to the `intervalaverage` function:

```{r}
no2_by_person <- intervalintersect(
  x = no2, 
  y = address_history, 
  interval_vars = c('start_date', 'end_date'), 
  group_vars = 'location_id'
)  
head(no2_by_person)
```

To understand what's happening here, let's look at person 5.

This person moved from location 40 to location 41 in November of 1996.

```{r}
address_history[person_id == 5]
```

If you look at the result of `intervalintersect`, you'll see that person
5 has two `no2` values for 1996, and that each one is associated with
the part of the year when they lived where that value applied.

```{r}
no2_by_person[person_id == 5 & year(start) %in% 1995:1997]
```

Now that we have a set of NO~2~ values associated with the `person_id`
rather than `location_id`, we can use `intervalaverage` to find the
average for any period of time. That could be a period that is unique to each participant. 
Or it could be consistent for all people in the data set, such as a yearly average
(where each year is a time-weighted average of the places the person
lived that year). 

We will work through both, and introduce one more function along the way. 

### Average Over the Year Prior to Study Enrollment 

The package includes another data.table with each person's enrollment date in our
study. We will use it to calculate the average exposure for each participant in the 
year before they joined our study. 

```{r}
data(enrollment)
setDT(enrollment)
head(enrollment)
```

Again, the `y` argument of `intervalaverage` tells it what date range to
average for every group in the values data set (`x`). So we first need to create a 
`data.table` that contains the beginning and end of our desired averaging period for each participant from the `enrollment`
data set. We'll define the "year prior" to enrollment in our study as the 365 days preceding the enrollment date
(i.e. the period from 365 days before enrollment to 1 day before enrollment). 

```{r}
pre_enrollment <- enrollment[ , .(person_id, start = enroll_date - 365L, end = enroll_date - 1L)]
head(pre_enrollment)
```

_You may notice I've used `start` and `end` here rather than `start_date` and `end_date` as I did previously. 
This is because of a quirk of `intervalintersect` where it returns `start` and `end` as the interval column names
regardless of the input names, and in the next step, the interval column names need to match. This will be fixed
in future versions of the package._ 

Now we simply apply `intervalaverage`, again allowing for 20% missingness in `no2` values:

```{r}
no2_pre_enrollment <- intervalaverage(
  x = no2_by_person, 
  y = pre_enrollment, 
  interval_vars = c('start', 'end'), 
  group_vars = 'person_id',
  value_vars = 'no2',
  required_percentage = 80
)  
head(no2_pre_enrollment[, .(person_id, start, end, no2)])
```

### Annual Averages

Finally, let's calculate annual (calendar-year) averages for 2000-2004. We'll need to build a `y` argument 
for the `intervalaverage` function that has a start date, end date for each year. However, even though 
the periods we are averaging over are the same for each participant, we *also* need to repeat each
year once for every participant. This is because  because `y` must contain the same grouping variable(s) as `x`. 

This gives us an opportunity to try a convenient helper function in `intervalaverage`: `CJ.dt`. This is a more
convenient version of `data.table::CJ` (`CJ` as in "Cross Join") that can take `data.tables` as arguments instead of just vectors. 
By cross joining our unique `person_id`s with the years 2000-2004, we can construct a `y` argument that will work 
in the `intervalaverage` function. 

```{r}
# Build a data.table with start and end dates for each year
years <- data.table(year = 2000:2004) 
years[ , start := as.IDate(paste(year, "01", "01", sep = "-"))]
years[ ,   end := as.IDate(paste(year, "12", "31", sep = "-"))]
years[]
```

```{r}
# Cross join with the unique person_ids in our data
years <- CJ.dt(years, unique(address_history[ , .(person_id)]))
head(years[order(year, person_id)])
```

Now we can apply the `intervalaverage` function again to get a yearly average for each person. Recall that the 
NO~2~ data is also annual, so the only rows that will actually be averaged will be those for years where a person
moved (since a person will have had an NO~2~ value for each address they lived in that year). 

```{r}
no2_annual <- intervalaverage(
  x = no2_by_person, 
  y = years, 
  interval_vars = c('start', 'end'), 
  group_vars = 'person_id',
  value_vars = 'no2',
  required_percentage = 80
)  
head(no2_annual[, .(person_id, start, end, no2)])
```


## Further Reading

For a more detailed and advanced tutorial of how to use the `intervalaverage` package, see `vignette("intervalaverage-advanced")`. 

For information about the inner workings of the package (intended for maintainers), see `vignette("intervalaverage-technicaloverview")`.

